Semi-Supervised Factored Logistic Regression for High-Dimensional Neuroimaging Data

### 2015-06-04 paper_nips/paper.tex
Pass on the end

\paragraph{Optimization.}
The common objective was optimized
by gradient descent in the SS-FLogReg parameters.
%
The required gradients were obtained by using the chain rule to
backpropagate error derivatives.
We chose the \textit{rmsprop} solver \cite{rmsprop},
a refinement of stochastic gradient descent.
\textit{Rmsprop} dictates an adaptive learning rate
for each model parameter by
scaled gradients from a running average.
The batch size was set to $100$
(much expected redundancy in $X_{rest}$ and $X_{task}$),
matrix parameters were initalized by Gaussian random values multiplied
by $0.004$ (i.e., gain), and
bias parameters were initalized to zero.
%
The normalization factor and the update rule for $\theta$
are given by
% XXX Michael: It is crucial, but nowhere mentioned, that the optimization is
% XXX done in an alternating manner. The whole structure relies on this, by
% XXX construction it is vulnerable to optimizing too much towards labels or
% XXX rest at a time. I also identify this as one of the major weak points
% XXX of the method, so it needs to be stated in a way that is not easily
% XXX attackable.
% XXX Maybe we could also talk less about RMSprop.
%
% XXX Danilo: I think we are not alternating, Xrest and Xtask go in
% simultaneously at every bach in every epoch
\begin{eqnarray}
  \begin{split}
    \mathbf{v^{(t+1)}} &= \rho \mathbf{v^{(t)}} + (1 - \rho)\left(\nabla_{\theta} f(x^{(t)}, y^{(t)}, \theta^{(t)})\right)^2
% XXX: I have changed the first equation, please check
%v^{(t+1)} = \rho v^{(t+1)} + (1 - \rho)˜|\nabla f
\\
\theta^{(t+1)} &= \theta^{(t)} + \alpha \frac{\nabla_{\theta} f(x^{(t)}, y^{(t)}, \theta^{(t)})}{\sqrt{\mathbf{v^{(t+1)}} + \epsilon}},
  \end{split}
\end{eqnarray}
% XXX Michael: Why, in the formula, do we have two different notations for
% XXX partial derivatives? \partial and \nabla
% XXX Michael: Probably remove the explanations following and just state values
% XXX: Bertrand: what is f ?? Danilo: replaced f by more detailed notation
where $f$ is the loss function computed on a minibatch sample at timestep $t$,
$\alpha$ is the learning rate ($0.00001$),
$\epsilon$ a global damping factor ($10^{-6}$)
and
$\rho$ the decay rate ($0.9$ to deemphasize the magnitude of the gradient).
%
Note that we have also experimented with other solvers
(stochastic gradient descent, adadelta, and adagrad) but found that
\textit{rmsprop} converged faster and with
similar or higher generalization performance.

@@

We first tested for an advantage of combined unsupervised decomposition

@@

We benchmarked against

@@

These explorations attest to the advantage of directly searching for

@@

We made several noteworthy observations (Table \ref{table_two}).
%
First, the purely supervised estimator ($\lambda=1$) achieved in no
instance the best accuracy, precision, recall or f1 scores on HCP data.
Classification by factored logistic regression is therefore facilitated by
imposing structure from the unlabeled rest data.
%
Second, the higher the number of latent components $n$,
the higher the out-of-study performance with small choices for $\lambda$.
This suggests that the presence of more rest-data-inspired hidden components
results in more effective feature representations in unrelated task data.
%
Third, for $n=20$ and $100$ (but not $5$) the purely rest-data-trained
decomposition matrix ($\lambda=0$) resulted in
noninferior out-of-study performance
of 77,0\% and 79,2\%,
respectively (Table \ref{table_two}).
This confirms that guiding model learning by task-unrelated structure
extracts features of general relevance
beyond the supervised problem at hand.

@@

\paragraph{Individual effects of dimensionality reduction and rest data.}
We first quantified the gain of introducing a bottleneck layer
disregarding the autoencoder.
To this end, ordinary logstic regression was juxtaposed
with FLogReg  at $\lambda=1$ (no autoencoder).
For this experiment, we increased the difficulty of the classification problem
by including data from all 38 HCP tasks.
Indeed, increased class separability in component space,
as compared to voxel space, entails differences in
generalization performance of
$\approx{17\%}$ (Figure \ref{fig_dimred}).
%
Notably,
the cognitive tasks on reward and punishment processing
are among the least predicted with ordinary but well predicted with
factored logistic regression
(tasks 3 and 4 in Figure \ref{fig_dimred}).
These experimental conditions have indeed been reported to exhibit
highly similar neural activity patterns in
GLM analyses of that dataset \cite{barch2013}.
Consequently, also local activity differences
(in the striatum and visual cortex in this case) can
be successfully captured by brain-network modelling.
%
\begin{figure}
\begin{centering}
\includegraphics[width=1.00\textwidth]{figures/accuracies.pdf}
\end{centering}
\vspace{-0.5cm}
\caption{\textbf{Effect of bottleneck in a 38-task classificaton problem}
Depicts the f1 score for each of 38 class predictions (chance is  $2,6\%$)
sorted from lowest to highest.
Multinomial logistic regression operating in voxel space (\textit{blue bars})
was compared to 
factored logistic regression operating in $20$
(\textit{left plot}) and $100$ (\textit{right plot})
latent modes (\textit{grey bars}).
Autoencoder or rest data were not used for these analyses
($\lambda=1$).
Ordinary logistic regression yielded 77,7\% accuracy out of sample, while
factored logistic regreesion scored at 94,4\% ($n=20$) and 94,2\% ($n=100$).
Hence, compressing the voxel data into a component space for classification
achieves higher task separability.
}
\label{fig_dimred}
\end{figure}

@@

\section{Discussion and Conclusion}
% Conceptual recap+gain of current paper
Using the flexibility of factored models,
we learn the low-dimensional representation from high-dimensional
voxel brain space that is most important for
prediction of cognitive task sets.
%
From a machine-learning perspective,
factorization of the logistic regression weights
can be viewed as transforming a
``multi-class classification problem''
into a ``multi-task learning problem''.
%
The higher generalization accuracy and feature recovery, comparing to
ordinary logistic regression, hold potential
for adoption in various neuroimaging analyses.
Besides increased performance, these models are more interpretable by
automatically learning a mapping to and from a brain-network space.
%
This domain-specifc learning algorithm
encourages departure from the artificial and statistically
less attractive voxel space.
Neurobiologically,
brain activity underlying defined mental operations
can be explained by linear combinations of the main activity
patterns.
%
That is,
fMRI data probably concentrate near
a low-dimensional manifold of
characteristic brain networks combinations.
Extracting fundamental building blocks of brain organization might
facilitate the quest for the cognitive primitives of
human thought.
%
We hope that these first steps stimulate development towards
powerful semi-supervised representation extraction
in systems neuroscience.

>>>>

\paragraph{Optimization.}
The common objective was optimized
by gradient descent in the SS-FLogReg parameters.
% XXX you haven't defined FLogReg
%
The required gradients were obtained by using the chain rule to
backpropagate error derivatives.
We chose the \textit{rmsprop} solver \cite{rmsprop},
a refinement of stochastic gradient descent.
\textit{Rmsprop} dictates an adaptive learning rate
for each model parameter by
scaled gradients from a running average.
The batch size was set to $100$
(much expected redundancy in $X_{rest}$ and $X_{task}$),
matrix parameters were initalized by Gaussian random values multiplied
by $0.004$ (i.e., gain), and
bias parameters were initalized to zero.
%
The normalization factor and the update rule for $\theta$
are given by
% XXX Michael: It is crucial, but nowhere mentioned, that the optimization is
% XXX done in an alternating manner. The whole structure relies on this, by
% XXX construction it is vulnerable to optimizing too much towards labels or
% XXX rest at a time. I also identify this as one of the major weak points
% XXX of the method, so it needs to be stated in a way that is not easily
% XXX attackable.
% XXX Maybe we could also talk less about RMSprop.
%
% XXX Danilo: I think we are not alternating, Xrest and Xtask go in
% simultaneously at every bach in every epoch
\begin{eqnarray}
  \begin{split}
    \mathbf{v^{(t+1)}} &= \rho \mathbf{v^{(t)}} + (1 - \rho)\left(\nabla_{\theta} f(x^{(t)}, y^{(t)}, \theta^{(t)})\right)^2
% XXX: I have changed the first equation, please check
%v^{(t+1)} = \rho v^{(t+1)} + (1 - \rho)˜|\nabla f
\\
\theta^{(t+1)} &= \theta^{(t)} + \alpha \frac{\nabla_{\theta} f(x^{(t)}, y^{(t)}, \theta^{(t)})}{\sqrt{\mathbf{v^{(t+1)}} + \epsilon}},
  \end{split}
\end{eqnarray}
% XXX Michael: Why, in the formula, do we have two different notations for
% XXX partial derivatives? \partial and \nabla
% XXX Michael: Probably remove the explanations following and just state values
% XXX: Bertrand: what is f ?? Danilo: replaced f by more detailed notation
where $f$ is the loss function computed on a minibatch sample at timestep $t$,
$\alpha$ is the learning rate ($0.00001$),
$\epsilon$ a global damping factor ($10^{-6}$)
and
$\rho$ the decay rate ($0.9$ to deemphasize the magnitude of the gradient).
%
Note that we have also experimented with other solvers
(stochastic gradient descent, adadelta, and adagrad) but found that
\textit{rmsprop} converged faster and with
similar or higher generalization performance.

@@

We first tested whether there is a substantial advantage in combining unsupervised decomposition

@@

We benchmarked our approach against

@@

This experiment establishes the advantage of directly searching for

@@

We made several noteworthy observations (Table \ref{table_two}).
%
First, the purely supervised estimator ($\lambda=1$) achieved in no
instance the best accuracy, precision, recall or f1 scores on HCP data.
Classification by factored logistic regression is therefore facilitated by
imposing structure from the unlabeled rest data.
%
Second, the higher the number of latent components $n$,
the higher the out-of-study performance with small values of $\lambda$.
This suggests that the presence of more rest-data-inspired hidden components
results in more effective feature representations in unrelated task data.
%
Third, for $n=20$ and $100$ (but not $5$) the purely rest-data-trained
decomposition matrix ($\lambda=0$) resulted in
noninferior out-of-study performance
of 77,0\% and 79,2\%,
respectively (Table \ref{table_two}).
This confirms that guiding model learning by task-unrelated structure
extracts features of general relevance
beyond the supervised problem at hand.

@@

\paragraph{Individual effects of dimensionality reduction and rest data.}
We first quantified the impact of introducing a bottleneck layer
that disregards the autoencoder.
To this end, ordinary logstic regression was juxtaposed
with FLogReg  at $\lambda=1$ (no autoencoder).
For this experiment, we increased the difficulty of the classification problem
by including data from all 38 HCP tasks.
Indeed, increased class separability in component space,
as compared to voxel space, entails differences in
generalization performance of
$\approx{17\%}$ (Figure \ref{fig_dimred}).
%
Notably,
the cognitive tasks on reward and punishment processing
are among the least predicted with ordinary but well predicted with
factored logistic regression
(tasks 3 and 4 in Figure \ref{fig_dimred}).
These experimental conditions have indeed been reported to exhibit
highly similar neural activity patterns in
GLM analyses of that dataset \cite{barch2013}.
Consequently, also local activity differences
(in the striatum and visual cortex in this case) can
be successfully captured by brain-network modelling.
%
\begin{figure}
\begin{centering}
\includegraphics[width=1.00\textwidth]{figures/accuracies.pdf}
\end{centering}
\vspace{-0.5cm}
\caption{\textbf{Effect of bottleneck in a 38-task classificaton problem}
Depicts the f1 score for each of 38 class predictions (chance is  $2,6\%$)
sorted from lowest to highest.
Multinomial logistic regression operating in voxel space (\textit{blue bars})
was compared to 
factored logistic regression operating in $20$
(\textit{left plot}) and $100$ (\textit{right plot})
latent modes (\textit{grey bars}).
Autoencoder or rest data were not used for these analyses
($\lambda=1$).
Ordinary logistic regression yielded 77,7\% accuracy out of sample, while
factored logistic regreesion scored at 94,4\% ($n=20$) and 94,2\% ($n=100$).
Hence, compressing the voxel data into a component space for classification
achieves higher task separability.
}
\label{fig_dimred}
\end{figure}

@@

\section{Discussion and Conclusion}
% Conceptual recap+gain of current paper
Using the flexibility of factored models,
we learn the low-dimensional representation from high-dimensional
voxel brain space that is most important for
prediction of cognitive task sets.
%
From a machine-learning perspective,
factorization of the logistic regression weights
can be viewed as transforming a
``multi-class classification problem''
into a ``multi-task learning problem''.
%
The higher generalization accuracy and feature recovery, comparing to
ordinary logistic regression, hold potential
for adoption in various neuroimaging analyses.
Besides increased performance, these models are more interpretable by
automatically learning a mapping to and from a brain-network space.
%
This domain-specific learning algorithm
encourages departure from the artificial and statistically
less attractive voxel space.
Neurobiologically,
brain activity underlying defined mental operations
can be explained by linear combinations of the main activity
patterns.
%
That is,
fMRI data probably concentrate near
a low-dimensional manifold of
characteristic brain networks combinations.
Extracting fundamental building blocks of brain organization might
facilitate the quest for the cognitive primitives of
human thought.
%
We hope that these first steps stimulate development towards
powerful semi-supervised representation extraction
in systems neuroscience.
